### 2025-08-05 causal-inference.qmd
20250708 - track chunk execution times
20250722 - chapter intro
20250805 - updates

@@

A causal effect is an effect in which the level of one input variable results in different levels on the outcome variable.
For instance, for a given player, the extent to which consumption of sports drink influences their performance reflects the difference between how well the player would perform if they consume sports drink compared to how well they would perform if they do not consume sports drink.

@@

The first criterion for establishing causality involves temporal precedence.
In order for a cause to influence an effect, the cause must occur before the effect.
For instance, if sports drink consumption influences player performance, the sports drink consumption (that is presumed to influence performance) must occur prior to the performance improvement.
Establishing the first criterion eliminates the possibility that the association between the purported cause and effect reflects reverse causation.
*Reverse causation* occurs when the purported effect is actually the cause of the purported cause, rather than the other way around.
For instance, if sports drink consumption occurs only once, and it occurs only before and not after performance, then we have ruled out the possibility of reverse causation (i.e., that better performance causes players to consume sports drink).

@@

The third criterion involves ruling out alternative reasons why the purported cause and effect may be related.
As noted in @sec-correlationCausality, there are four reasons why `X` may be correlated with `Y`.
If we meet the first criterion of causality, we have removed the possibility that `Y` causes `X` (i.e., reverse causality).
To meet the third criterion of causality, we need to remove the possibility that the association reflects a third variable ([confound](#sec-causalDiagramConfounding)) that influences both the cause and effect, and we need to remove the possibility that the association is spurious—the possibility that the association between the purported cause and effect is due to random chance.

@@

In general, to meet the third criterion of causality, one must consider the counterfactual—what would have happened "counter to the fact".
A *counterfactual* is what would have happened in the hypothetical scenario that the cause did not occur [i.e., what would have happened in the absence of the cause; @Shadish2002].
When engaging in causal inference, it is important to consider what would have happened if the hypothetical cause had actually not occurred.
It is like an alternate universe where the cause did not happen.
For instance, consider that we conduct an experiment to randomly assign some players to consume a sports drink before a game and other players to drink only water.
In this case, our treatment/intervention group is the group of players that consumed a sports drink.
The control group is the group players that drank only water.
Now, consider that the players in the treatment group outperform the players in the the control group in their football game.
In such a study, we observe what *did happen* when players received a treatment.
The counterfactual is knowledge of what *would have happened* to those same players if they simultaneously had not received treatment [@Shadish2002].
The true causal effect, then, is the difference between what did happen and what would have happened.
However, we cannot observe a counterfactual.
This is the fundamental problem of causal inference—we only observe one potential world/outcome/universe.
That is, we do not know for sure what would have happened to the players who received treatment if those same players had actually not received treatment.
We have a control group, but the control group does not have the same players as the intervention group, and it is impossible for a person to simultaneously receive and not receive treatment.

@@

As described in @sec-experiment, [experimental designs](#sec-experiment) are designs in which participants are randomly assigned to one or more levels of the [independent variable](#sec-experiment) to observe its effects on the [dependent variable](#sec-experiment).
[Experimental designs](#sec-experiment) provide the strongest tests of causality because they can rule out reverse causation and third variables.
For instance, by manipulating sports drink consumption before the player performs, they can eliminate the possibility that reverse causation explains the effect of the [independent variable](#sec-experiment) on the [dependent variable](#sec-experiment).
Second, through randomly assigning players to consume or not consume sports drink, this holds everything else constant (so long as the groups are evenly distributed according to other factors, such as their age, weight, etc.) and thus removes the possibility that third variable [confounds](#sec-causalDiagramConfounding) explain the effect of the [independent variable](#sec-experiment) on the [dependent variable](#sec-experiment).

@@

Another design feature of [longitudinal designs](#sec-causalInferenceLongitudinal) that can lead to greater [internal validity](#sec-internalValidity) is the use of within-subject analyses.
Between-subject analyses, might examine, for instanc, whether players who consume more sports drink perform better on average compared to players who consume less sports drink.
However, there are other between-person differences that could explain any observed between-subject associations between sports drink consumption and players performance.
Another approach could be to apply within-subject analyses.
For instance, you could examine whether, within the same individual, if a player consumes a sports drink, do they perform better compared to games in which they did not consume a sports drink.
When we control for prior levels of the outcome in the prediction, we are evaluating whether the predictor is associated with witin-person *change* in the outcome.
Predicting within-person change provides stronger evidence consistent with causality because it uses the individual as their own control and controls for many time-invariant [confounds](#sec-causalDiagramConfounding) (i.e., [confounds](#sec-causalDiagramConfounding) that do not change across time).
However, predicting within-person change does not, by itself, control for time-varying [confounds](#sec-causalDiagramConfounding).
So, it can also be useful to control for time-varying [confounds](#sec-causalDiagramConfounding), such as by use of [control variables](#sec-causalInferenceControlVariables).

@@

However, in observational studies, the groups may differ on the [confounding factors](#sec-causalDiagramConfounding); people may choose the level of the they receive.
Thus, we need to attempt to control for the [confounds](#sec-causalDiagramConfounding); however, in an observational study, we can only potentially adjust for the observed [confounds](#sec-causalDiagramConfounding), unlike in an [experiment](#sec-causalInferenceExperiment) in which we can control for both observed and unobserved [confounds](#sec-causalDiagramConfounding).

@@

For instance, researchers may compare monozygotic twins versus dizygotic twins in some outcome—a so-called "twin study".
It is assumed that the trait/outcome is attributable to genetic factors to the extent that the monozygotic twins (who share 100% of their DNA) are more similar in the trait or outcome compared to the dizygotic twins (who share on average 50% of their DNA).
Alternatively, researchers could compare full siblings versus half-siblings, or they could compare full siblings versus first cousins.

@@

In a causal model, an *ancestor* is a variable that influences another variable, either directly or indirectly via another variable [@Rohrer2018].
A *descendant* is a variable that is influenced by another variable [@Rohrer2018].
In general, one should not control for descendants of the outcome variable, because doing so could eliminate the apparent effect of a legitimate cause on the outcome variable [@Digitale2022].
Moreover, as described above, when trying to understand the total causal effect of a predictor variable on an outcome variable, one should not control for descendants of the predictor variable that are also antecedents of the outcome variable (i.e., mediators of the effect of the predictor variable on the outcome variable) [@Digitale2022].

@@

There are three criteria for establishing causality: 1) the cause precedes the effect.
2) The cause is related to the effect.
3) There are no other alternative explanations for the effect apart from the cause.
In general, it is important to be aware of the counterfactual and to consider what would have happened if the supposed cause had not occurred.
Various experimental and quasi-experimental designs and approaches can be leveraged to more closely approximate causal inferences.
[Longitudinal designs](#sec-causalInferenceLongitudinal), [within-subject analyses](#sec-causalInferenceWithinSubject), [inclusion of control variables](#sec-causalInferenceControlVariables), and [genetically informed designs](#sec-causalInferenceGeneticallyInformed) are all quasi-experimental designs that afford the researcher greater control over some possible third variable [confounds](#sec-causalDiagramConfounding).
Causal diagrams can be a useful tool for identifying the proper variables to control for (and those not to control for).
When [confounding](#sec-causalDiagramConfounding) exists, it is important to control for the [confound(s)](#sec-causalDiagramConfounding).
It is important not to control for mediators when interested in the total effect of the [predictor variable](#sec-correlationalStudy) on the [outcome variable](#sec-correlationalStudy).
In addition, it is important not to control for [descendants](#sec-ancestorsDescendants) of the [outcome variable](#sec-correlationalStudy).
When there is a [collision](#sec-colliderBias), it is important not to control for the [collider](#sec-colliderBias) (unless the [collider](#sec-colliderBias) is also a [confound](#sec-causalDiagramConfounding)).

>>>>

This chapter provides an overview of principles of causal inference and [causal diagrams](#sec-causalDiagrams).

@@

A causal effect is the difference in an outcome that is directly attributable to different levels or values of an input variable.
It reflects the difference in the outcome that would be observed for the same unit under two or more different levels of the input variable, holding all other relevant factors constant.
For instance, for a given player, the extent to which consumption of sports drink influences their performance reflects the difference between how well the player would perform if they consume sports drink compared to how well they would perform if they do not consume sports drink.

@@

The first criterion for establishing causality involves temporal precedence.
In order for a cause to influence an effect, the cause must occur before the effect.
For instance, if sports drink consumption influences player performance, the sports drink consumption (that is presumed to influence performance) must occur prior to the performance improvement.
Establishing the first criterion eliminates the possibility that the association between the purported cause and effect reflects reverse causation.
*Reverse causation* occurs when the purported effect is actually the cause of the purported cause, rather than the other way around.
For instance, if sports drink consumption occurs only once, and it occurs only before and not after performance, then we have ruled out the possibility of reverse causation (i.e., that better performance causes players to consume sports drink).
To help establish the first criterion, [experiments](#sec-causalInferenceExperiment) and [longitudinal designs](#sec-causalInferenceLongitudinal) are useful.

@@

The third criterion involves ruling out alternative reasons why the purported cause and effect may be related.
As noted in @sec-correlationCausality, there are five reasons why `X` may be correlated with `Y`.
If we meet the first criterion of causality, we have removed the possibility that `Y` causes `X` (i.e., reverse causality).
To meet the third criterion of causality, we need to remove the possibility that the association reflects a third variable ([confound](#sec-causalDiagramConfounding)) that influences both the cause and effect, and we need to remove the possibility that the association is spurious—the possibility that the association between the purported cause and effect is due to random chance.

@@

In general, to meet the third criterion of causality, one must consider the counterfactual—what would have happened "counter to the fact".
A *counterfactual* is what would have happened in the hypothetical scenario that the cause did not occur [i.e., what would have happened in the absence of the cause; @Shadish2002].
When engaging in causal inference, it is important to consider what would have happened if the hypothetical cause had actually not occurred.
It is like an alternate universe where the cause did not happen.
For instance, consider that we conduct an experiment to randomly assign some players to consume a sports drink before a game and other players to drink only water.
In this case, our treatment/intervention group is the group of players that consumed a sports drink.
The control group is the group players that drank only water.
Now, consider that the players in the treatment group outperform the players in the control group in their football game.
In such a study, we observe what *did happen* when players received a treatment.
The counterfactual is knowledge of what *would have happened* to those same players if they simultaneously had not received treatment [@Shadish2002].
The true causal effect, then, is the difference between what did happen and what would have happened.
However, we cannot observe a counterfactual.
This is the fundamental problem of causal inference—we only observe one potential world/outcome/universe.
That is, we do not know for sure what would have happened to the players who received treatment if those same players had actually not received treatment.
We have a control group, but the control group does not have the same players as the intervention group, and it is impossible for a person to simultaneously receive and not receive treatment.

@@

As described in @sec-experiment, [experimental designs](#sec-experiment) are designs in which participants are randomly assigned to one or more levels of the [independent variable](#sec-experiment) to observe its effects on the [dependent variable](#sec-experiment).
[Experimental designs](#sec-experiment) provide the strongest tests of causality because they can rule out reverse causation and third variables.
For instance, by manipulating sports drink consumption before the player performs, they can eliminate the possibility that reverse causation explains the effect of the [independent variable](#sec-experiment) on the [dependent variable](#sec-experiment).
Second, through randomly assigning players to consume or not consume sports drink, this holds everything else constant (so long as the groups are evenly distributed according to other factors, such as their age, weight, etc.) and thus removes the possibility that third variable [confounds](#sec-causalDiagramConfounding) explain the effect of the [independent variable](#sec-experiment) on the [dependent variable](#sec-experiment).
However, not everything is able to, or ethical to, be manipulated.
In addition, as described in @sec-internalExternalValidityTradeoffs, experimental designs tend to have lower capacity for [external validity](#sec-externalValidity) than [correlational designs](#sec-correlationalStudy)—that is, compared to findings from [correlational designs](#sec-correlationalStudy), the findings from experimental designs tend to have lower capacity to be consistent with how things play out in the real world.

@@

Another design feature of [longitudinal designs](#sec-causalInferenceLongitudinal) that can lead to greater [internal validity](#sec-internalValidity) is the use of within-subject analyses.
Between-subject analyses, might examine, for instance, whether players who consume more sports drink perform better on average compared to players who consume less sports drink.
However, there are other between-person differences that could explain any observed between-subject associations between sports drink consumption and players performance.
Another approach could be to apply within-subject analyses.
For instance, you could examine whether, within the same individual, if a player consumes a sports drink, do they perform better compared to games in which they did not consume a sports drink.
When we control for prior levels of the outcome in the prediction, we are evaluating whether the predictor is associated with within-person *change* in the outcome.
Predicting within-person change provides stronger evidence consistent with causality because it uses the individual as their own control and controls for many time-invariant [confounds](#sec-causalDiagramConfounding) (i.e., [confounds](#sec-causalDiagramConfounding) that do not change across time).
However, predicting within-person change does not, by itself, control for time-varying [confounds](#sec-causalDiagramConfounding).
So, it can also be useful to control for time-varying [confounds](#sec-causalDiagramConfounding), such as by use of [control variables](#sec-causalInferenceControlVariables).

@@

However, in observational studies, the groups may differ on the [confounding factors](#sec-causalDiagramConfounding); for instance, due to other factors (e.g., personality, beliefs, and attitudes), people may choose (or self-select into) the level of the variable (e.g., sport drink consumption) they receive.
Thus, we need to attempt to control for the [confounds](#sec-causalDiagramConfounding); however, in an observational study, we can only potentially adjust for the observed [confounds](#sec-causalDiagramConfounding), unlike in an [experiment](#sec-causalInferenceExperiment) in which we can control for both observed and unobserved [confounds](#sec-causalDiagramConfounding).

@@

Genetically informed designs evaluate the extent to which individuals who share greater genetic relatedness are more similar in a characteristic compared to individuals who share less genetic relatedness.
For instance, researchers may compare monozygotic twins versus dizygotic twins in some outcome—a so-called "twin study".
It is assumed that the trait/outcome is attributable to genetic factors to the extent that the monozygotic twins (who share 100% of their DNA) are more similar in the trait or outcome compared to the dizygotic twins (who share on average 50% of their DNA).
Alternatively, researchers could compare full siblings versus half-siblings, or they could compare full siblings versus first cousins.

@@

In a causal model, an *ancestor* is a variable that influences another variable, either directly or indirectly via another variable [@Rohrer2018].
A *descendant* is a variable that is influenced by another variable [@Rohrer2018].
In general, one should not control for descendants of the outcome variable, because doing so could eliminate the apparent effect of a legitimate cause on the outcome variable [@Digitale2022].
Moreover, as described above, when trying to understand the total causal effect of a predictor variable on an outcome variable, one should not control for descendants of the predictor variable that are also ancestors of the outcome variable (i.e., mediators of the effect of the predictor variable on the outcome variable) [@Digitale2022].

@@

There are three criteria for establishing causality: 1) the cause precedes the effect.
2) The cause is related to the effect.
3) There are no other alternative explanations for the effect apart from the cause.
In general, it is important to be aware of the counterfactual and to consider what would have happened if the supposed cause had not occurred.
Various experimental and quasi-experimental designs and approaches can be leveraged to more closely approximate causal inferences.
[Longitudinal designs](#sec-causalInferenceLongitudinal), [within-subject analyses](#sec-causalInferenceWithinSubject), [inclusion of control variables](#sec-causalInferenceControlVariables), and [genetically informed designs](#sec-causalInferenceGeneticallyInformed) are all quasi-experimental designs that afford the researcher greater control over some possible third variable [confounds](#sec-causalDiagramConfounding).
Causal diagrams can be a useful tool for identifying the proper variables to control for (and those not to control for).
In general, the only variables you should control for are [confound(s)](#sec-causalDiagramConfounding).
When [confounding](#sec-causalDiagramConfounding) exists, it is important to control for the [confound(s)](#sec-causalDiagramConfounding).
It is important not to control for mediators when interested in the total effect of the [predictor variable](#sec-correlationalStudy) on the [outcome variable](#sec-correlationalStudy).
In addition, it is important not to control for [descendants](#sec-ancestorsDescendants) of the [outcome variable](#sec-correlationalStudy).
When there is a [collision](#sec-colliderBias), it is important not to control for the [collider](#sec-colliderBias) (unless the [collider](#sec-colliderBias) is also a [confound](#sec-causalDiagramConfounding)).
